---
title: "Data Science Lab Research Project"
subtitle: "[Choose sub-title yourself]"
author:
  - name: "[Denis Lebedev]"
    st. number: "4826973"
    email: "d.lebedev@students.uu.nl"
  - name: "Tomi ____"
    st. number: "0000000"
    email: "t._____@students.uu.nl"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: paged
    code_folding: show
    highlight: tango

params:
  course: "Data Science Lab (ECB3DSL), Academic year 2024/2025"
  team: "Group 1"
  supervisor: "TBD"
  word_count: "TBD"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "90%"
)

# Load packages you expect to use (add/remove as needed)
library(tidyverse)
library(broom)
library(modelr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(janitor)
library(skimr)
```

# Title page

**Course:** `r params$course`  
**Team:** `r params$team`  

**Students:**  
- [Student Name 1, Student Number 1, Student Email 1]  
- [Student Name 2, Student Number 2, Student Email 2]  

**Supervisor(s):** `r params$supervisor`  

**Word count:** `r params$word_count`  

---

# Abstract

*(At most 100 words. Motivate the research question, what you do, what you find, and why it matters. Elevator pitch style.)*

[Write abstract here.]

---

# Statement of Originality

We, the aforementioned students, herewith declare to have written this document and that we are responsible for the content of it. We declare that the text and the work presented in this document is original and that no sources other than those mentioned in the text and its references have been used in creating it.

Utrecht University School of Economics is responsible solely for the supervision of completion of the work, not for the content.

---

# Disclosure Statement (only if applicable)

In this project, we have made use of the following Generative AI tools:  
- [List the tools]

We have used these tools for:  
- [Short statement of what you used the tool(s) for]

---

# Division of Work

We, the aforementioned students, herewith declare that we have divided the work on this project and this project paper as stated in the following table:

| Section and Content | Name and Student Number |
|---|---|
| 1 Introduction | All students |
| 2 Data | [Student Name & Number 1] |
| 3 Descriptive Analyses | [Student Name & Number 2] |
| 4 Empirical Approach | [All students / Assigned] |
| 5 Results | [Student Name & Number 3] |
| 6 Conclusion | All students |

---

# Signatures

*(Insert scans of signatures here, plus student name and number.)*

- [Signature image / placeholder] — [Student Name, Student Number]  
- [Signature image / placeholder] — [Student Name, Student Number]  
- [Signature image / placeholder] — [Student Name, Student Number]  

---

# 1 Introduction

[Discuss: overall topic; specific research question; why it matters (societal + scientific debate, link to literature); short overview of main findings; brief overview of methodology.]

## 1.1 Research question and motivation

[Write here.]

## 1.2 Contribution and related literature (brief)

[Write here.]

## 1.3 Preview of approach and findings

[Write here.]

---

# 2 Data

[Describe the dataset and preparations: variable definitions, cleaning, exclusions, transformations — only what is needed to understand the project.]

## 2.1 Data source and sample construction

[Write here.]

## 2.2 Variables and definitions

[Write here.]

## 2.3 Data cleaning and preprocessing

```{r data-cleaning}
df <- read_csv("UCI_Credit_Card.csv")

df <- df %>% clean_names()
```
### Summary

```{r}
skim(df)
```

### Number \ Share of Default Payments

```{r}
df %>%
  count(default_payment_next_month) %>%
  mutate(share = n / sum(n))

df %>%
  count(sex) %>%
  mutate(share = n/sum(n))
```

---

# 3 Descriptive Analyses

[Provide descriptive analyses needed to understand key features of the data relevant for your research question.]

## 3.1 Summary statistics

```{r summary-stats}
# TODO: Summary table
# df_clean %>%
#   summarise(across(where(is.numeric), list(mean = mean, sd = sd), na.rm = TRUE))
```

## 3.2 Visual exploration

```{r descriptive-plots}
# TODO: Example plot
# ggplot(df_clean, aes(x = var_x)) +
#   geom_histogram(bins = 30)
```

## 3.3 Correlations / relationships (if relevant)

```{r correlations}
# TODO: Example correlation matrix (numeric vars only)
# df_num <- df_clean %>% select(where(is.numeric))
# cor_mat <- cor(df_num, use = "pairwise.complete.obs")
# knitr::kable(round(cor_mat, 3))
```

---

# 4 Empirical Approach

[Explain the empirical approach. This depends on your project (prediction vs causal inference, supervised vs unsupervised, etc.).]

## 4.1 Model / method overview

[Write here.]

## 4.2 Identification / assumptions (if relevant)

[Write here.]

## 4.3 Implementation details

```{r empirical-approach}
# TODO: Specify train/test split, cross-validation, model specs, etc.
# set.seed(123)
# split <- initial_split(df_clean, prop = 0.8)
# train <- training(split)
# test  <- testing(split)
```

---

# 5 Results

[Show, explain, and discuss your results.]

## 5.1 Main results

```{r main-results}
# TODO: Fit model(s) and show main outputs
# model <- lm(y ~ x1 + x2, data = train)
# summary(model)
# tidy(model)
```

## 5.2 Model evaluation / robustness (if relevant)

```{r evaluation}
# TODO: Example prediction + metrics
# preds <- predict(model, newdata = test)
# rmse <- sqrt(mean((test$y - preds)^2, na.rm = TRUE))
# rmse
```

## 5.3 Discussion

[Interpret results in plain language and relate back to the research question and literature.]

---

# 6 Conclusion

[Conclude: What do you find? How does that contribute to the research question? How does that contribute to the relevant scientific and societal debate? What remains open?]

## 6.1 Key takeaways

[Write here.]

## 6.2 Limitations

[Write here.]

## 6.3 Future research

[Write here.]

---

# References

*(Use your preferred style; below is a placeholder format.)*

- Last Name 1, First Name 1, and Last Name 2, First Name 2 (year): “Title”, *Journal Name*, Volume(Issue): PageX–PageY.

```{r references-helper, eval=FALSE}
# Optional: if you use BibTeX
# 1) create references.bib
# 2) add to YAML:
# bibliography: references.bib
# csl: apa.csl
```

---

# Appendix (optional)

[Optional appendix for robustness checks, extra figures/tables, implementation details, etc.]

## A.1 Additional tables

```{r appendix-tables}
# TODO
```

## A.2 Additional figures

```{r appendix-figures}
# TODO
```
