---
title: "Data Science Lab Research Project"
author:
- name: Denis Lebedev
  st. number: '4826973'
  email: d.lebedev@students.uu.nl
- name: Tamas Kiss
  st. number: '4944216'
  email: t.kiss@students.uu.nl
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: paged
    code_folding: show
    highlight: tango
subtitle: ''
params:
  course: Data Science Lab (ECB3DSL), Academic year 2025/2026
  team: Group 2
  supervisor: Matías Golman
  word_count: TBD
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "90%"
)

# Load packages you expect to use (add/remove as needed)
library(tidyverse)
library(broom)
library(modelr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(janitor)
library(skimr)
library(corrplot)
library(reshape2)
library(modelsummary)
```

# Title page

**Course:** `r params$course`  
**Team:** `r params$team`  

**Students:**  
- [Denis Lebedev, 4826973, d.lebedev@students.uu.nl]  
- [Tamas Kiss, 4944216, t.kiss@students.uu.nl]  

**Supervisor(s):** `r params$supervisor`  

**Word count:** `r params$word_count`  

---

# Abstract

*(At most 100 words. Motivate the research question, what you do, what you find, and why it matters. Elevator pitch style.)*

[Write abstract here.]

---

# Statement of Originality

We, the aforementioned students, herewith declare to have written this document and that we are responsible for the content of it. We declare that the text and the work presented in this document is original and that no sources other than those mentioned in the text and its references have been used in creating it.

Utrecht University School of Economics is responsible solely for the supervision of completion of the work, not for the content.

---

# Disclosure Statement (only if applicable)

In this project, we have made use of the following Generative AI tools:  
- Open AI's "Chat GPT"

We have used these tools for:  
- R code generating based on prompts, explicitly stating the logic, theory and purpose of the code

---

# Division of Work

We, the aforementioned students, herewith declare that we have divided the work on this project and this project paper as stated in the following table:

| Section and Content | Name and Student Number |
|---|---|
| 1 Introduction | All students |
| 2 Data | [Student Name & Number 2] |
| 3 Descriptive Analyses | [Student Name & Number 2] |
| 4 Empirical Approach | [All students / Assigned] |
| 5 Results | [Student Name & Number 3] |
| 6 Conclusion | All students |

---

# Signatures

*(Insert scans of signatures here, plus student name and number.)*

- [Signature image / placeholder] — [Student Name, Student Number]  
- [Signature image / placeholder] — [Student Name, Student Number]  
- [Signature image / placeholder] — [Student Name, Student Number]  

---

# 1 Introduction

[Discuss: overall topic; specific research question; why it matters (societal + scientific debate, link to literature); short overview of main findings; brief overview of methodology.]

## 1.1 Research question and motivation

Research Question Ideas:

1) Do such variables as sex, age, education and marital status help predict the probability of default?
  - age can mean more experience and more savings -> less probability of default
  - marriage can help mitigate the default due to common savings
  - education can mean more financial awareness -> less probability of default

2) Do previous loan payments help predict probability of default?
  - diligent payments -> signal of low prob-ty of default
  - or do people pay whenever it is convenient, and past payment are not good indicators?
  
3) Does the loan amount serve as a default indicator?
  - more debt -> higher default rate?
  - loan amount categories?

Methodology ideas:
 - supervised learning
 - no specific models until we look at the data
    (for 1 - regression, lasso\ridge
     for 2 - same
     for 3 - could be RDD)

## 1.2 Contribution and related literature (brief)

[Write here.]

## 1.3 Preview of approach and findings

[Write here.]

---

# 2 Data

[Describe the dataset and preparations: variable definitions, cleaning, exclusions, transformations — only what is needed to understand the project.]

## 2.1 Data source and sample construction

[Write here.]

## 2.2 Variables and definitions

Complete list of variables:

```
ID: ID of each client
LIMIT_BAL: Amount of given credit in NT dollars (includes individual and family/supplementary credit)
SEX: Gender (1=male, 2=female)
EDUCATION: (1=graduate school, 2=university, 3=high school, 4=others, 5=unknown, 6=unknown)
MARRIAGE: Marital status (1=married, 2=single, 3=others)
AGE: Age in years
PAY_0: Repayment status in September, 2005 (-1=pay duly, 1=payment delay for one month, 2=payment delay for two months, … 8=payment delay for eight months, 9=payment delay for nine months and above)
PAY_2: Repayment status in August, 2005 (scale same as above)
PAY_3: Repayment status in July, 2005 (scale same as above)
PAY_4: Repayment status in June, 2005 (scale same as above)
PAY_5: Repayment status in May, 2005 (scale same as above)
PAY_6: Repayment status in April, 2005 (scale same as above)
BILL_AMT1: Amount of bill statement in September, 2005 (NT dollar)
BILL_AMT2: Amount of bill statement in August, 2005 (NT dollar)
BILL_AMT3: Amount of bill statement in July, 2005 (NT dollar)
BILL_AMT4: Amount of bill statement in June, 2005 (NT dollar)
BILL_AMT5: Amount of bill statement in May, 2005 (NT dollar)
BILL_AMT6: Amount of bill statement in April, 2005 (NT dollar)
PAY_AMT1: Amount of previous payment in September, 2005 (NT dollar)
PAY_AMT2: Amount of previous payment in August, 2005 (NT dollar)
PAY_AMT3: Amount of previous payment in July, 2005 (NT dollar)
PAY_AMT4: Amount of previous payment in June, 2005 (NT dollar)
PAY_AMT5: Amount of previous payment in May, 2005 (NT dollar)
PAY_AMT6: Amount of previous payment in April, 2005 (NT dollar)
default.payment.next.month: Default payment (1=yes, 0=no)
```

## 2.3 Data cleaning and preprocessing

```{r data-cleaning}
df <- read_csv("UCI_Credit_Card.csv")

df <- df %>% clean_names()

df <- df %>% select(-id) #remove ID, not predictive
colSums(is.na(df)) #check for NA values
df <- df %>% distinct() #check for duplicates

#converting numerical encoding
df <- df %>%
  mutate(sex = factor(sex,
                      levels = c(1, 2),
                      labels = c("Male", "Female")))

#cleaning education
df <- df %>%
  mutate(education = case_when(
    education == 1 ~ "Graduate School",
    education == 2 ~ "University",
    education == 3 ~ "High School",
    education == 4 ~ "Other",
    TRUE ~ "Unknown"
  )) %>%
  mutate(education = factor(education))

#cleaning marital status
df <- df %>%
  mutate(marriage = case_when(
    marriage == 1 ~ "Married",
    marriage == 2 ~ "Single",
    marriage == 3 ~ "Other",
    TRUE ~ "Unknown"
  )) %>%
  mutate(marriage = factor(marriage))

#cleaning default status
df <- df %>%
  rename(default = default_payment_next_month) %>%
  mutate(default = factor(default,
                          levels = c(0,1),
                          labels = c("No Default", "Default")))
```
### Summary

```{r}
skim(df)
```

### Number \ Share of Default Payments

```{r}
df %>%
  count(default) %>%
  mutate(share = n / sum(n))

df %>%
  count(sex) %>%
  mutate(share = n/sum(n))
```

---

# 3 Descriptive Analyses

[Provide descriptive analyses needed to understand key features of the data relevant for your research question.]

## 3.1 Summary statistics

###Summary table
```{r summary-stats}
df %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd), na.rm = TRUE))
```

## 3.2 Visual exploration
```{r}
library(tidyverse)
library(scales)

theme_ds <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold", size = base_size + 4),
      plot.subtitle = element_text(size = base_size + 1, color = "grey30"),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      plot.caption = element_text(color = "grey40")
    )
}

default_rate_by <- function(data, var) {
  data %>%
    group_by({{ var }}) %>%
    summarise(
      n = n(),
      default_rate = mean(default == "Default"),
      .groups = "drop"
    ) %>%
    mutate({{ var }} := fct_reorder({{ var }}, default_rate))
}
```

###Default rate by sex
```{r descriptive-plots}
overall_rate <- mean(df$default == "Default")

p_sex <- default_rate_by(df, sex) %>%
  ggplot(aes(x = sex, y = default_rate)) +
  geom_col(width = 0.4, fill = "#2C7FB8") +
  
  # overall benchmark line
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             linewidth = 0.8,
             color = "red") +
  annotate("text",
           x = Inf,
           y = overall_rate,
           label = paste0("Overall: ", scales::percent(overall_rate, accuracy = 0.1)),
           hjust = 1.1,
           vjust = -0.5,
           color = "red",
           fontface = "bold",
           size = 3.5) +
  
  geom_text(aes(label = scales::percent(default_rate, accuracy = 0.1)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.1))
  ) +
  
  labs(
    title = "Default rate by sex",
    subtitle = "Compared to overall default rate",
    x = NULL,
    y = "Default rate"
  ) +
  theme_ds()

p_sex
```

###Default rate by education
```{r}
p_edu <- default_rate_by(df, education) %>%
  ggplot(aes(x = education, y = default_rate)) +
  geom_col(width = 0.7, fill = "#2C7FB8") +
  geom_text(aes(label = scales::percent(default_rate, accuracy = 0.1)),
            vjust = -0.4, fontface = "bold", size = 3.3) +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             linewidth = 0.8,
             color = "red") +
  annotate("text",
           x = Inf,
           y = overall_rate,
           label = paste0("Overall: ", scales::percent(overall_rate, accuracy = 0.1)),
           hjust = 0.2,
           vjust = -0.5,
           color = "red",
           fontface = "bold",
           size = 3.5) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Default rate by education",
    subtitle = "Ordered from lowest to highest default rate",
    x = NULL, y = "Default rate"
  ) +
  coord_cartesian(clip = "off") +
  theme_ds() +
theme(
  axis.text.x = element_text(),
  plot.margin = margin(5.5, 70, 5.5, 5.5)
)
p_edu
```

###Default rate by marital status
```{r}
p_mar <- default_rate_by(df, marriage) %>%
  ggplot(aes(x = marriage, y = default_rate)) +
  geom_col(width = 0.6, fill = "#2C7FB8") +
  geom_text(aes(label = percent(default_rate, accuracy = 0.1)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             linewidth = 0.8,
             color = "red") +
  annotate("text",
           x = Inf,
           y = overall_rate,
           label = paste0("Overall: ", scales::percent(overall_rate, accuracy = 0.1)),
           hjust = 0,2,
           vjust = -0.5,
           color = "red",
           fontface = "bold",
           size = 3.5) +
  scale_y_continuous(labels = percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Default rate by marital status",
    subtitle = "Share of clients who defaulted next month",
    x = NULL, y = "Default rate"
  ) +
  coord_cartesian(clip = "off") +
  theme_ds() +
theme(
  axis.text.x = element_text(),
  plot.margin = margin(5.5, 70, 5.5, 5.5)
)
p_mar
```

###Default rate by age group
```{r}
df <- df %>%
  mutate(age_group = cut(age,
                         breaks = c(20, 25, 30, 35, 40, 45, 50, 60, 80),
                         right = FALSE))

age_summary <- df %>%
  group_by(age_group) %>%
  summarise(
    default_rate = mean(default == "Default"),
    .groups = "drop"
  ) %>%
  filter(!is.na(age_group))

p_age <- age_summary %>%
  ggplot(aes(x = age_group, y = default_rate)) +
  geom_col(width = 0.75, fill = "#2C7FB8") +
  geom_text(aes(label = paste0(percent(default_rate, accuracy = 0.1))),
            vjust = -0.5, size = 3.1, fontface = "bold") +
   geom_hline(yintercept = overall_rate,
             linetype = "dashed",
             linewidth = 0.8,
             color = "red") +
  annotate("text",
           x = Inf,
           y = overall_rate,
           label = paste0("Overall: ", scales::percent(overall_rate, accuracy = 0.1)),
           hjust = 0,2,
           vjust = -0.5,
           color = "red",
           fontface = "bold",
           size = 3.5) +
  scale_y_continuous(labels = percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Default rate by age group",
    subtitle = "Default rate with sample size per group",
    x = NULL, y = "Default rate"
  ) +
  coord_cartesian(clip = "off") +
  theme_ds() +
theme(
  axis.text.x = element_text(),
  plot.margin = margin(5.5, 70, 5.5, 5.5)
)

p_age
```

### Further visual exploration of suggested RQs 2 and 3

```{r}

# ----------------------------
# Dynamic Path: past repayments
# ----------------------------

pay_vars <- c("pay_0","pay_2","pay_3","pay_4","pay_5","pay_6")

df_dyn <- df %>%
  mutate(
    worst_delay = pmax(!!!rlang::syms(pay_vars), na.rm = TRUE),
    months_late = rowSums(across(all_of(pay_vars), ~ .x >= 1), na.rm = TRUE)
  )

# (1) Default rate by worst observed delay status
dyn_worst <- df_dyn %>%
  group_by(worst_delay) %>%
  summarise(
    n = n(),
    default_rate = mean(default == "Default"),
    .groups = "drop"
  ) %>%
  mutate(worst_delay = factor(worst_delay, levels = sort(unique(worst_delay))))

p_dyn1 <- dyn_worst %>%
  ggplot(aes(x = worst_delay, y = default_rate)) +
  geom_col(width = 0.75, fill = "#2C7FB8") +
  geom_text(aes(label = scales::percent(default_rate, accuracy = 0.1)),
            vjust = -0.4, size = 3.1, fontface = "bold") +
  geom_hline(yintercept = overall_rate, linetype = "dashed", linewidth = 0.8, color = "red") +
  scale_y_continuous(labels = percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Dynamic path: default rate by worst repayment status (pay_*)",
    subtitle = "Worst delay observed across the last 6 months",
    x = "Worst repayment status (higher = more delayed)", y = "Default rate"
  ) +
  coord_cartesian(clip = "off") +
  theme_ds() +
  theme(plot.margin = margin(5.5, 30, 5.5, 5.5))

p_dyn1

# (2) Default rate by number of months with delay (>= 1)
dyn_late <- df_dyn %>%
  group_by(months_late) %>%
  summarise(
    n = n(),
    default_rate = mean(default == "Default"),
    .groups = "drop"
  )

p_dyn2 <- dyn_late %>%
  ggplot(aes(x = months_late, y = default_rate)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_text(aes(label = scales::percent(default_rate, accuracy = 0.1)),
            vjust = -0.8, size = 3.1, fontface = "bold") +
  geom_hline(yintercept = overall_rate, linetype = "dashed", linewidth = 0.8, color = "red") +
  scale_y_continuous(labels = percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
  scale_x_continuous(breaks = sort(unique(dyn_late$months_late))) +
  labs(
    title = "Dynamic path: default rate vs. frequency of delayed payments",
    subtitle = "Months late = count of pay_* >= 1 (across 6 months)",
    x = "Number of months late", y = "Default rate"
  ) +
  coord_cartesian(clip = "off") +
  theme_ds() +
  theme(plot.margin = margin(5.5, 30, 5.5, 5.5))

p_dyn2


# ----------------------------
# Volume Path: loan amount (credit limit)
# ----------------------------

df_vol <- df %>%
  filter(!is.na(limit_bal)) %>%
  mutate(limit_decile = ntile(limit_bal, 10))

vol_dec <- df_vol %>%
  group_by(limit_decile) %>%
  summarise(
    n = n(),
    default_rate = mean(default == "Default"),
    mean_limit = mean(limit_bal),
    .groups = "drop"
  )

# (1) Default rate by limit_bal decile
p_vol1 <- vol_dec %>%
  ggplot(aes(x = limit_decile, y = default_rate)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_text(aes(label = scales::percent(default_rate, accuracy = 0.1)),
            vjust = -0.8, size = 3.1, fontface = "bold") +
  geom_hline(yintercept = overall_rate, linetype = "dashed", linewidth = 0.8, color = "red") +
  scale_y_continuous(labels = percent, limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Volume path: default rate by credit limit (limit_bal) deciles",
    subtitle = "Decile 1 = lowest credit limits, Decile 10 = highest",
    x = "limit_bal decile", y = "Default rate"
  ) +
  coord_cartesian(clip = "off") +
  theme_ds() +
  theme(plot.margin = margin(5.5, 30, 5.5, 5.5))

p_vol1

# (2) Distribution view: limit_bal by default status
p_vol2 <- df_vol %>%
  mutate(default = factor(default, levels = c("No Default", "Default"))) %>%
  ggplot(aes(x = default, y = limit_bal)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Volume path: credit limit distribution by default status",
    subtitle = "Boxplot of limit_bal for Default vs No Default",
    x = NULL, y = "limit_bal"
  ) +
  theme_ds()

p_vol2
```



## 3.3 Correlations / relationships (if relevant)

###Correlation check
```{r}
pay_vars <- c("pay_0","pay_2","pay_3","pay_4","pay_5","pay_6")

df_corr <- df %>%
  mutate(
    # Dynamic path
    avg_repay_status = rowMeans(across(all_of(pay_vars)), na.rm = TRUE),

    # Outcome
    default_num = ifelse(default == "Default", 1, 0),

    # Sex: 1 = Male
    sex_male = ifelse(sex == "Male", 1, 0),

    # Education ordered (higher = more education)
    education_level = case_when(
      education == "Graduate School" ~ 3,
      education == "University" ~ 2,
      education == "High School" ~ 1,
      TRUE ~ 0   # Other / Unknown
    ),

    # Married dummy
    married = ifelse(marriage == "Married", 1, 0)
  )

corr_data <- df_corr %>%
  select(default_num, limit_bal, avg_repay_status, age, sex_male, education_level, married) %>%
  mutate(across(everything(), as.numeric))  # final safety

cor_matrix <- cor(corr_data) #, use = "pairwise.complete.obs"

cor_long <- melt(cor_matrix)

ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC", midpoint = 0) +
  theme_ds() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Matrix (Key Variables)", x = NULL, y = NULL)
```
###Starter model
```{r}
model1 <- glm(
  default_num ~ age + sex_male + education_level + married,
  data = df_corr,
  family = binomial
)

summary(model1)
```


```{r}
options(scipen = 999, digits = 3)

model1 <- glm(
  default_num ~ age + sex_male + education_level + married,
  data = df_corr,
  family = binomial
)

summary(model1)

model2 <- glm(
  default_num ~ age + sex_male + education_level + married + limit_bal,
  data = df_corr,
  family = binomial
)

summary(model2)

model3 <- glm(
  default_num ~ age + sex_male + education_level + married + limit_bal + avg_repay_status,
  data = df_corr,
  family = binomial
)

summary(model3)
```

---

# 4 Empirical Approach

[Explain the empirical approach. This depends on your project (prediction vs causal inference, supervised vs unsupervised, etc.).]

## 4.1 Model / method overview

[Write here.]

## 4.2 Identification / assumptions (if relevant)

[Write here.]

## 4.3 Implementation details

```{r empirical-approach}
# TODO: Specify train/test split, cross-validation, model specs, etc.
# set.seed(123)
# split <- initial_split(df_clean, prop = 0.8)
# train <- training(split)
# test  <- testing(split)
```

---

# 5 Results

[Show, explain, and discuss your results.]

## 5.1 Main results

```{r main-results}
# TODO: Fit model(s) and show main outputs
# model <- lm(y ~ x1 + x2, data = train)
# summary(model)
# tidy(model)
```

## 5.2 Model evaluation / robustness (if relevant)

```{r evaluation}
# TODO: Example prediction + metrics
# preds <- predict(model, newdata = test)
# rmse <- sqrt(mean((test$y - preds)^2, na.rm = TRUE))
# rmse
```

## 5.3 Discussion

[Interpret results in plain language and relate back to the research question and literature.]

---

# 6 Conclusion

[Conclude: What do you find? How does that contribute to the research question? How does that contribute to the relevant scientific and societal debate? What remains open?]

## 6.1 Key takeaways

[Write here.]

## 6.2 Limitations

[Write here.]

## 6.3 Future research

[Write here.]

---

# References

*(Use your preferred style; below is a placeholder format.)*

- Last Name 1, First Name 1, and Last Name 2, First Name 2 (year): “Title”, *Journal Name*, Volume(Issue): PageX–PageY.

```{r references-helper, eval=FALSE}
# Optional: if you use BibTeX
# 1) create references.bib
# 2) add to YAML:
# bibliography: references.bib
# csl: apa.csl
```

---

# Appendix (optional)

[Optional appendix for robustness checks, extra figures/tables, implementation details, etc.]

## A.1 Additional tables

```{r appendix-tables}
# TODO
```

## A.2 Additional figures

```{r appendix-figures}
# TODO
```
